---
title: "01_yield"
format: html
editor: visual
---

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(readxl)
library(tidymodels)
library(sf)
library(ggplot2)
```

# Training

## Meta data

```{r}
meta <- read_csv("../data/training/training_meta.csv")

meta
```

<<<<<<< HEAD

There are 43 distinct sites. However, MOH1 is replicated twice and TXH1 is replicated 3 times. 
=======
There are 43 distinct sites.

>>>>>>> e8e01d0bd91d9c321f16bb64fc20c1955c40368a
```{r}
meta %>% 
  distinct(site)
```

#GEH1 site seems to be far from the other data collected.

```{r}
meta %>% 
  st_as_sf( coords = c("longitude", "latitude"), crs = 4326) %>% 
  ggplot()+
  geom_sf()#+
  geom_sf_text(aes(label = site, size = 0.1))
```

Lets add base map of US to visualize the spatial data distribution

```{r warning=FALSE}
library(maps)

us_map <- st_as_sf(map("state", plot = F, fill = T))

meta %>% 
  st_as_sf( coords = c("longitude", "latitude"), crs = 4326) %>% 
  ggplot()+
  geom_sf(data = us_map, fill = "gray82", color = "white")+
  geom_sf(color = "red", size = 2)
```

GEH1 site does not lie within the US boundary. So, let's drop that for

```{r}
summary(meta)
```

## Soil data

```{r}
soil <- read_csv("../data/training/training_soil.csv") %>% 
  mutate(site = substr(site,1,4))

soil
```

```{r}
summary(soil)
```

## Trait data

```{r}
trait <- read_csv("../data/training/training_trait.csv") %>% 
  mutate(date_planted = as.Date(date_planted, format = "%m/%d/%y"),
         date_harvested = as.Date(date_harvested, format = "%m/%d/%y")) %>% 
  mutate(growing_l_d = date_harvested - date_planted) %>% 
  mutate(growing_l_d = substr(growing_l_d, 1,3)) %>%
  mutate(growing_l_d = as.numeric(growing_l_d))

trait
```

```{r}
summary(trait)
```

# Merge

```{r message=FALSE}
weather_df <- read_csv("../data/training/weathermonthsum_train.csv") 

df <- trait %>% 
  group_by(year, site) %>% 
  left_join(soil, relationship = "many-to-many") %>% 
  left_join(weather_df, relationship = "many-to-many")

summary(df)
```


```{r}

# wea_df <- read_csv("../data/training/fieldweatherdata_train.csv") %>% 
#   dplyr::select(-c(site_2:year_2)) %>% 
#   mutate(date = as.Date(paste0(yday,"/",year), format = "%j/%Y"))

```


unique site
```{r}
df %>%
  distinct(site) %>%
  arrange(site)
```

unique previous crops - do we need to worry about previous crops??
```{r}
df %>%
  distinct(previous_crop) %>%
  arrange(previous_crop)
```

Clean previous_crop column
```{r}
df <- df %>%
  mutate(previous_crop = case_when(
    is.na(previous_crop)                          ~ "unknown",
    str_detect(str_to_lower(previous_crop), "soy")     ~ "soybean",
    str_detect(str_to_lower(previous_crop), "corn")    ~ "corn",
    str_detect(str_to_lower(previous_crop), "wheat")   ~ "wheat",
    str_detect(str_to_lower(previous_crop), "grain")   ~ "wheat",
    str_detect(str_to_lower(previous_crop), "2019")    ~ "wheat",
    str_detect(str_to_lower(previous_crop), "cotton")  ~ "cotton",
    str_detect(str_to_lower(previous_crop), "sorghum") ~ "sorghum",
    str_detect(str_to_lower(previous_crop), "peanut")  ~ "peanut",
    str_detect(str_to_lower(previous_crop), "fallow")  ~ "fallow",
    TRUE                                               ~ "other"
  ))


df %>% count(previous_crop, sort = TRUE)
```


Export joined file
```{r export}
write_csv(df, "../data/training/df_joined.csv")
```



```{r}
df %>% 
  filter(year == "2014")
```
