---
title: "03_elevatr"
format: html
editor: visual
---

```{r message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidymodels)

library("terra")
library("elevatr")
library(sf)
```

# Sites

Extract locations from the meta file to download the parameter of interest.

```{r}
meta <- read_csv("../data/training/training_meta.csv") %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  filter(!site == "GEH1")

```

# Elevatr

The elevatr package provides access to elevation data from AWS Open Data Terrain Tiles and the Open Topography Global datasets API for raster digital elevation models.

Source:  [[https://github.com/usepa/elevatr]].

```{r}
elev <- get_elev_point(meta, src = "epqs") #point scale takes a lot of time and internal server error is seen with code "500".
```

There are some NA is previous dataframe. So, let's extract the sites where the elevation data is not available.

```{r}
ele_na <- elev %>% 
  filter(is.na(elevation)) %>% 
  dplyr::select(site, geometry) # 55 sites donot have elevation data
```

Second lot of elevation data extraction.

```{r}
elev_02 <- get_elev_point(ele_na, src = "epqs")
```

# Merge 

Merge all the sites with complete set of information on elevation.

```{r}
elev_all <- elev %>% 
  st_join(elev_02) %>% 
  mutate(elevation = dplyr::case_when(is.na(elevation.x) ~ elevation.y,TRUE ~ elevation.x)) %>%
  select(-elevation.x, -elevation.y, -elev_units.x, -elev_units.y, -site.y) %>% 
  rename(site = site.x,
         elevation_m = elevation)
```

# Save the output

```{r}
# write_sf(elev_all,"../data/training/03_elevation_m.geojson")
```


