---
title: "XgBoost"
format: html
editor: visual
---


```{r}
#shared_lib_path <- "/work/crss8030/instructor_data/shared_R_libs"

#class_packages <- c("tidymodels", "tidyverse", "vip", "ranger", "finetune", "parsnip", "reticulate", "xgboost", "doParallel","nloptr", "lme4", "here")

#install.packages(class_packages, lib = shared_lib_path)
```

# Locate the installed libraries (important step while running the script in cluster - sapelo2)

```{r}
# .libPaths(c("/work/crss8030/instructor_data/shared_R_libs", .libPaths()))
```


## Step 1: Load Libraries
```{r}

 #.libPaths()
```

We begin by loading the required libraries:

Here we have **2 new packages** that you havenot seen before in the class :: here and quarto.

```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage

library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
library(lme4)         # The matrix package and other mathmatical computation for xgboost
library(here)         # Gives the absolute path starting from Rproj, works well with different scripts (R, .rmd or .qmd)
library(quarto)       # helps to export the qmd script to R script
#library(caret)       # Other great library for Machine Learning 

here::here()
```

# Load the data set

-   Setting a seed ensures reproducibility.
-   Splitting data helps validate generalizability by preventing the model from seeing the test data during training.

```{r weather}
# read using here package
weather <- read_csv(here("data", "training", "wea_month_sum_yield.csv")) %>%
  select(-1) %>%
  mutate(year = as.factor(year)) %>% 
  drop_na()

summary(weather)
```

-   We begin by splitting the dataset into **training** and **testing** sets.
-   We use **stratified sampling** to ensure that the distribution of the target variable is maintained in both sets.
-   A random **seed** is set to ensure reproducibility of the split.
-   This step ensures that the model is validated on unseen data, helping assess its **generalizability**.

> ⚠️ Note: Since tree-based models (like XGBoost and Random Forest) are not affected by the scale of the features, **normalization or standardization is not necessary**

# ML workflow

## 1. Pre-processing

```{r weather_split}
set.seed(931735) # Setting seed to get reproducible results 

weather_split <- initial_split(
  weather, 
  prop = .7, # proption of split same as previous codes
  strata = yield_mg_ha  # Stratify by target variable
  )

weather_split
```

### a. Data split

For data split, let's use **70% training / 30% testing**.

```{r weather_train}
weather_train <- training(weather_split)  # 70% of data
weather_train #This is your traing data frame
```

```{r weather_test}
weather_test <- testing(weather_split)    # 30% of data
weather_test
```

### b. Distribution of target variable

```{r distribution}
density_plot <- ggplot() +
  geom_density(data = weather_train, 
               aes(x = yield_mg_ha),
               color = "red") +
  geom_density(data = weather_test, 
               aes(x = yield_mg_ha),
               color = "blue") 

# lets print the undergoing process in the console, (important when running in the terminal later on)
cat(paste0("Saving...... density plot of train and test data"))

ggsave(plot = density_plot, 
       path = here("output", "png"),
       filename = "density_plot_test_train.png",
       height = 6,
       width = 9,
       dpi = 600)
  
```

### c. Data processing with recipe

Before training, we need to perform some processing steps, like

-   **removing unimportant variables**
-   dropping NAs
-   removing columns with single value
-   Updating the role of some useful columns (like year and location)
-   others?

For that, we'll create a **recipe** of these processing steps.

This recipe will then be applied now to the **train data**, and easily applied to the **test data** when we bring it back at the end.

Creating a recipe is an easy way to port your processing steps for other data sets without needing to repeat code, and also only considering the data it is being applied to.

You can find all available recipe step options here: https://tidymodels.github.io/recipes/reference/index.html

Different model types require different processing steps.\
Let's check what steps are required for an elastic net model (linear_reg). We can search for that in this link: https://www.tmwr.org/pre-proc-table #Don't need for numeric data

> Differently from elastic net, variables do not need to be normalized in XgBoost, so we'll skip this step.

```{r weather_recipe}
cat("\nCreate recipe for data preprocessing\n")

weather_recipe <- recipe(yield_mg_ha ~ ., data = weather_train) %>% # Remove identifier columns and months not in growing season
  step_rm(
    matches("Jan|Feb|Mar|Apr|Oct|Nov|Dec")  # Remove non-growing season months
  ) %>%
  # updating the role of the 2 columns as ID, so they arenot used as predictors
  update_role(c(year, site), new_role = "ID") %>% 
  step_other(hybrid, threshold = 500/nrow(weather_train)) %>%  # groups rare hybrids into "other"
  step_dummy(hybrid, one_hot = TRUE)          # converts hybrid to numeric columns

weather_recipe
```

```{r weather_prep}
# Prep the recipe to estimate any required statistics
weather_prep <- weather_recipe %>% 
  prep()

# Examine preprocessing steps
weather_prep
```


```{r}
weather_prep %>% 
  summary()
```


## 2. Training

### a. Model specification

First, let's specify:\
- the **type of model** we want to train\
- which **engine** we want to use\
- which **mode** we want to use

> XgBoost can be run with either numerical (regression) or categorical (classification) explanatory variable. Therefore, we have the need to specify the mode here.

XgBoost **hyperparameters**:

-   **`trees`**: The number of boosting rounds (i.e., how many trees will be added sequentially).\
-   **`tree_depth`**: Controls how deep each individual tree can grow. Deeper trees can capture more complex interactions but also increase the risk of overfitting.\
-   **`min_n`**: Minimum number of observations required in a node for it to be split. Acts as a regularization tool to prevent overly specific splits.\
-   **`learn_rate`**: Also known as `eta`, it controls how much each additional tree contributes to the overall model. Smaller values make the model more stable but require more trees.

```{r}
# helps to see all the available hyperparameter we can tune for a model
show_model_info("boost_tree")


# helps to see all the models that we can run with the package parsnip
get_from_env("models")
```


```{r xgb_spec}
xgb_spec <- #Specifying XgBoost as our model type, asking to tune the hyperparameters
  boost_tree(
   # Total number of boosting iterations
    trees = tune(),
         # Maximum depth of each tree
    tree_depth = tune(),
             # Minimum samples required to split a node
    min_n = tune(),
        # Step size shrinkage for each boosting step
    learn_rate = tune()
      ) %>%
        #specify engine 
  set_engine("xgboost") %>%
       # Set to mode
  set_mode("regression")
  
xgb_spec

```

### b. Cross-validation setup

-   We use 10-fold cross-validation to evaluate model performance during tuning:

-   Here we are adding 2 new concepts of CV technique:

1.  Leave one year out (important when we are doing temporal prediction)

2.  Leave one state out (important when we are doing geographical prediction)

```{r}
set.seed(235) #34549

resampling_foldcv <- vfold_cv(weather_train, # Create 10-fold cross-validation resampling object from training data
                              v = 10)

# # Create leave one year out cv object from the sampling data
# resampling_fold_loyo <- group_vfold_cv(weather_train,
#                                        group = year)
# 
# 
# # Create leave one location out cv object from the sampling data
# resampling_fold_loso <- group_vfold_cv(weather_train,
#                                        group = site)

```

### c. Hyperparameter grid with Latin Hypercube Sampling

We use Latin hypercube sampling to generate a diverse grid of hyperparameter combinations:

```{r }
set.seed(12345)

xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  size = 20
)

xgb_grid
```

```{r}
ggplot(data = xgb_grid,
       aes(x = tree_depth, 
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate),
                 size = trees),
             alpha = .5,
             show.legend = FALSE)
```

## 3. Detecting total cores

-   This help us to detect the total number of available cores and then do a parallel processing on each of them

```{r}
# Detect Cores for Slurm (Cluster - Sapelo2)

# lets look into our current environment to see the available cores
n_cores <- as.numeric(Sys.getenv("SLURM_CPUS_ON_NODE"))

if (is.na(n_cores)) n_cores <- parallel::detectCores() - 1

# Start the Cluster
cl <- makePSOCKcluster(n_cores)

registerDoParallel(cl)

cat(paste0("\nFound and registered ", n_cores, " cores to work with\n"))

```

## 4. Model Tuning

To explore this extensive grid, we are going to use a method called **ANOVA Racing**.

It works by:

1.  Evaluating all of the candidate models, but only for a few resamples.

2.  Determining which candidates have a low probability of being selected.

3.  Eliminate poor candidates.

4.  Repeat with next resample (until no more resamples remain).

This can result in fitting a small number of models. It is not an iterative search; it is an adaptive grid search.

```{r xgb_grid_result}

set.seed(76544)

# Create the list of CV techniques so we can loop through them
cv_list <- list(
  vfold = resampling_foldcv
  # year = resampling_fold_loyo,
  # location = resampling_fold_loso
)

# Create a empty list to store the results from the loop
results <- list()

# Create our loop
for (i in seq_along(cv_list)) {
  
  name <- names(cv_list)[i]
  
  results[[name]] <- tune_race_anova(object = xgb_spec,
                                     preprocessor = weather_recipe,
                                     resamples = cv_list[[i]],
                                     grid = xgb_grid,
                                     control = control_race(save_pred = TRUE,
                                                            parallel_over = "everything"))
}



stopCluster(cl)
results
```

Let's check the race results

```{r}
library(dplyr)

# Create a dataframe structure from the list obtained after running the loop
results_df <-  tibble(method = names(results),
                      diff_cv = results) %>%
  # Collect the metrices for each CV techniques using map function
  mutate(metrices = map2(diff_cv, method,
                         ~.x %>%
                           collect_metrics()%>%
                           mutate(method = .y, .before = "trees")))

results_df

# bind all the metrices together so to select the best performing one
all_metrices <- do.call(bind_rows, results_df$metrices)

all_metrices
```

```{r}
# Automating to pull the best method out of 3 we ran

best_method <- all_metrices %>%
  filter(.metric == "rmse") %>%
  slice_min(mean, n = 1) %>%
  pull(method)
  

# Getting the metrice (hyperparameter values of the best performing CV)
best_cv_object <- results_df %>%
  filter(method == best_method) %>%
  pull(diff_cv) %>% 
  first()
  
```

Notice how some combinations were dropped off early due to poor fit metrics on some initial resamples.

## 4. Select Best Models

We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:

```{r}
# Best RMSE
best_rmse <- best_cv_object %>% 
      select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse
```

Here we use all three methods which we learn in this class for R2.

```{r}
# Based on greatest R2
best_r2 <- best_cv_object %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```

## Compare and Finalize Model

```{r comparing values}
best_rmse %>% 
  bind_rows(best_rmse, 
            best_r2) %>%
  dplyr::select(source, everything())
```

## 5. Final Specification

```{r final_spec_fit}
final_spec <- boost_tree(
  trees = best_r2$trees,           # Number of boosting rounds (trees)
  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
  min_n = best_r2$min_n,           # Minimum number of samples to split a node
  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```

## 6. Final Fit and Predictions

## Validation

```{r final_fit}
set.seed(10)
final_fit <- last_fit(final_spec,
                weather_recipe,
                split = weather_split)

final_fit %>%
  collect_predictions()
```

```{r}
final_fit %>% collect_metrics()
```

## 7. Evaluate on Test Set

```{r final_fit_metrics}
test_met <- final_fit %>%
  collect_metrics() %>% 
  mutate(estimate = round(.estimate, 3))
```



## 8. Evaluate on Training Set

```{r}
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rmse(yield_mg_ha, .pred)
```

```{r}
# R2
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rsq(yield_mg_ha, .pred)
```


```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(0, 25)) +
  scale_y_continuous(limits = c(0, 25)) 
```

## 9. Predicted vs Observed Plot (publication ready)

```{r}
publication_ready <- final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha,
             y = .pred)) +
  # Changing the shape of the points to 
  geom_point(aes(fill = yield_mg_ha), shape = 21, alpha = 0.7, show.legend = F) +
  # Using colorblind friendly color scale 
  scale_fill_viridis_c(option = "H") +
  
  # Changing the color and linetype of abline
  geom_abline(color = "red", linetype = 2) +
  
  geom_smooth(method = "lm") +
  labs(x = "Observed Yield (Mg/Ha)",
       y = "Predicted Yield (Mg/Ha)")+
  # Annotate R-sq and RMSE value in the plot
  annotate("label", x = Inf, y = -Inf,
           label = paste0("R-sq:", test_met$estimate[[2]], "\nRMSE: ", test_met$estimate[[1]]),
           hjust = 1, vjust = -0.5) +
  
  scale_x_continuous(limits = c(0, 25)) +
  scale_y_continuous(limits = c(0, 25)) +
  
  # Changing the theme of the plot
    theme(panel.background = element_rect(fill = "gray82"))


cat(paste0("\nSaving...... publication ready plot for test dataset\n"))

ggsave(plot = publication_ready, 
       path = here("output", "png"),
       filename = "model_perf_test_data.png",
       height = 6,
       width = 9,
       dpi = 600)
  
```

## 10. Variable Importance

```{r final_spec}
vip <- final_spec %>%
  fit(yield_mg_ha ~ .,
         data = bake(weather_prep, weather_train) %>%
        select(-year, -site)) %>% 
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable,
             fill = Importance)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)+
  theme(panel.background  = element_rect(fill = "gray82"),
        panel.grid = element_blank())

cat(paste0("\nSaving...... publication ready plot for variable importance\n"))

ggsave(plot = vip, 
       path = here("output/png"),
       filename = "vip_test_data.png",
       height = 6,
       width = 9,
       dpi = 600)
```

```{r}
knitr::purl("04_XGBoost_model.qmd", output = "XGBoost_script.R", documentation = 0)
```
